---
title: "Unlocking single cell spatial omics analyses with SCDNEY - Demo"
author:
- name: Yue Cao^1,2,3^, Helen Fu^1,2,3^, Jean Yang^1,2,3^
  affiliation:
  - 1. Sydney Precision Data Science Centre, University of Sydney, Australia; \\
  - 2. School of Mathematics and Statistics, University of Sydney, Australia; \\
  - 3. Charles Perkins Centre, University of Sydney, Australia   
date: 1 August, 2023

params:
  evalc: FALSE   ## EDIT to TRUE when generating output, otherwise 'FALSE'
  show: 'hide'  ## EDIT to 'as.is' when generating Suggestions, otherwise 'hide'
output:
  html_document:
    css: https://use.fontawesome.com/releases/v5.0.6/css/all.css
    code_folding: hide
    fig_height: 12
    fig_width: 12
    toc: yes
    number_sections: true
    toc_depth: 3
    toc_float: yes
    self_contained: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message=FALSE, warning= FALSE)
```


```{r, eval=params$evalc}
## library(devtools)
## library(BiocManager)
library(SingleCellExperiment)
library(ggplot2)
library(scFeatures) ## devtools::install_github("SydneyBioX/scFeatures")
library(ClassifyR) ## BiocManager::install("ClassifyR", dependencies = TRUE)
library(ggthemes)
library(spicyR) ## BiocManager::install("spicyR")
library(dplyr)
library(limma)
library(plotly)
library(scattermore)
library(tidyr)
library(survival)
library(survminer)
library(spatstat)
##library(spatstat.core) ## install.packages("spatstat")
##library(spatstat.geom) 
library(scater)
library(scran)

theme_set(theme_classic())
```

# Overview

As single cell technology advances, the recent development of spatial omics allows us to examine the spatial organisation of cells within tissues in their native environment. This workshop will discuss the challenges and analytical focus associated with disease outcome prediction using multi-sample spatial datasets. We will also talk about general analytical strategies and the critical thinking questions that arise in the workflow.  

<br>
<div class="aimbox"> 
### <span class="fa-stack fa"><i class="fa fa-circle fa-stack-2x"></i><i class="fa fa-pencil-alt fa-stack-1x fa-inverse"></i></span> Preparation and assumed knowledge {-}
- Knowledge of R syntax
- Familiarity with the [SingleCellExperiment class](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) 
- Ability to install all required R packages, please check `sessionInfo` at the end of this document to ensure you are using the correct version. 

### <span class="fa-stack fa"><i class="fa fa-circle fa-stack-2x"></i><i class="fa fa-location-arrow fa-stack-1x fa-inverse"></i></span> Learning objectives {-}
- Understand and visualise spatial omics datasets     
- Explore various strategies for disease outcome prediction using spatial omics data      
- Understand and generate individual feature representations from a cell-level expression matrix 
- Develop appreciation on how to assess the performance of classification models   
- Perform disease outcome prediction using the feature representation and robust classification framework     
</div>

<br>

### Time outline {-}
 Structure for this 2-hour workshop:

 | Activity                     | Time    |
 |------------------------------|---------|
 | Introduction to spatial technologies    | 20m  |
 | Cell segmentation with deep learning (with BIDCell)   | 20m  |
 | Exploring spatial data    |  20m    |
 | Break (Q&A) | 10m  |
 | Extracting features from spatial data (with scFeatures)    |  30m    |
 | Performing disease outcome classification (with ClassifyR) |  20m    |

# Initial exploration and visualisation

#### Data and background {-} 
Imaging mass cytometry (IMC) is a new imaging technique that gathers spatial information to create images that show the distribution of different cell types and their associated protein expression patterns in tissue. 
In this demo, we examine IMC dataset taken from: 

Moldoveanu D *et. al.* Spatially mapping the immune landscape of melanoma using imaging mass cytometry. Science Immunology, Apr;7(70):eabi5072. doi: 10.1126/sciimmunol.abi5072. PMID: 35363543.

Here, the authors quantified the expression of 35 protein markers in two melanoma cohorts. Basic characteristics of the data objects:   

- The dataset contains 35 proteins and 112,497 cells.   
- The outcome is 16 non-responders (`Response` = No) and 14 responders (`Response` = Yes).    
- 17 patients received CTLA4 treatment, 8 patients received PD1, 5 patients received both.   
- From a range of tissue sources and primary sites.    

```{r eval=params$evalc}
load("../data/breastCancer.RData")
data_sce <- IMC
logcounts(data_sce) <- log2(assay(data_sce) + 1)

print("data format")
assay(data_sce)[1:5, 1:5]

##print("the object stores meta data (such as patient outcome information) about each cell")
##DT::datatable( data.frame(colData(data_sce))[1:5, ]  , options = list(scrollX = TRUE))
```

#### Aim {-} 
In this demo, we will examine one of the cohorts - pretreatment melanoma samples from 30 individuals with advanced disease who subsequently received ICI therapy. We will aim to identify features that has the potnetial to discriminate between non-responders (`Response` = No) and responders (`Response` = Yes).   

### Exploration 1: How complex is my data? {-} 
At the start of the exploration, it is often good to get a sense of the complexity of the data.  We usually do this by exploring the distribution of the outcomes and various variables in the patient's meta-data.  Here, we use cross-tabulation to examine the following variables:

- responder vs non-responders
- tissue source
- primary site

```{r eval=params$evalc, fig.height=5, fig.width=5}
print("Stage and death")
table(clinical$Breast.Surgery, clinical$Death, useNA = "ifany") 

print("Number of patients based on ER status")
table(clinical$ER.Status)

print("Number of patients based on Grade")
table(clinical$Grade)
```

### Exploration 2: How to visualise my data? {-} 
Typically in single-cell data analysis, we perform dimension reduction to project the high dimensional cell by gene matrix on to 2D space. This allows us to visualise various things of interest, such as distribution of cell types and disease outcomes. In this dataset, cells were classified in the following cell types based on their markers:   

- CD8+ cytotoxic T cells (Tc).   
- CD4+ T helper cells (Th).   
- macrophage and monocytes (macro.mono; CD68+ or CD14+).   
- B cells (B; CD20+), melanoma and melanocytes (melano; SOX10+ or S100+).   
- endothelial cells (CD31+).    
- cells that did not express any of these markers or with conflicting expression patterns were classified as “others”.      

*Note:* for single-cell RNA-seq with around 20,000 genes, we often need to perform some filtering (e.g. select highly variable genes) to reduce the number of features. Here, given we have less than 50 proteins, there is no need to pre-filter. That being said, we provide some sample code below (commented out) to demonstrate how to identify highly variable genes followed by UMAP visualisation in scRNA-seq data.

```{r eval=FALSE, include=TRUE}
# gene_var <- modelGeneVar(data_sce)
# hvgs <- getTopHVGs(gene_var, prop=0.1)
# data_sce <- runUMAP(data_sce, scale=TRUE,  subset_row = hvgs)
data_sce <- runUMAP(data_sce, scale=TRUE)
```
```{r eval=params$evalc,fig.height=4, fig.width=12}
a <- plotUMAP(data_sce, colour_by = "description")
b <- plotUMAP(data_sce, colour_by = "metabricId")
a + b
```

<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<i class="far fa-chart-bar fa-stack-1x fa-inverse"></i>
</span> <strong>Interactive Q&A:</strong>     

What can we learn from these illustrations? Is there anything interesting in the plot? Questions to consider include:    

- Q1: Does each cell type cluster together?
- Q2: When there is a large number of categories, are dimensionality reduction plots interpretable?     

</div>

### Exploration 3: Is there a spatial structure in my data? {-} 
The advantage with spatial omics is that we can examine the organisation of the cell types as it occurs on the tissue slide. Here, we visualise one of the slides from a patient. As an optional exercise, you can 

- permute the spatial coordinates 

to give a sense of what is random ordering. 

```{r fig.height=4, fig.width=6, eval=params$evalc}
one_sample <- data_sce[, data_sce$metabricId  == "MB-0263"]
one_sample  <- colData(one_sample)
one_sample <- data.frame(one_sample)

tableau_palette <- scale_colour_tableau() 
color_codes <- tableau_palette$palette(10)

a <- ggplot(one_sample, aes(x = Location_Center_X , y = Location_Center_Y, colour = description)) + geom_point(alpha=0.7) +  scale_colour_manual(values = c(color_codes, "lightgrey")) + ggtitle("Original slide")

```

```{r  fig.height=4, fig.width=12, eval=params$evalc}
print("Optional: Permute the cell type labels")
one_sample$description_permute <- sample(one_sample$description)
b <- ggplot(one_sample, aes(x = Location_Center_X , y = Location_Center_Y, colour =description_permute)) + geom_point(alpha=0.7) +  scale_colour_manual(values = c(color_codes, "lightgrey")) + ggtitle("Permute the cell type label")

print("Optional: Permute the spatial coordinate")
one_sample$Location_Center_X_permute <- sample(one_sample$Location_Center_X)
one_sample$Location_Center_Y_permute <- sample(one_sample$Location_Center_Y)
c <- ggplot(one_sample, aes(x = Location_Center_X_permute , y = Location_Center_Y_permute, colour = description)) + geom_point(alpha=0.7) +  scale_colour_manual(values = c(color_codes, "lightgrey")) + ggtitle("Permute the X, Y coordinate")
a + b + c
```

<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<i class="far fa-chart-bar fa-stack-1x fa-inverse"></i>
</span> <strong> Critical thinking: </strong>   

- Is there structure in the data real ? 
- What are additional strategies to generate a random distribution? 

</div>


#  Describing tissue microenvrionments and cellular neighbourhoods  

## Do cell type co-localise in specfic regions? 

We begin by examining how can we identify and visualise regions of tissue where spatial associations between cell-types are similar. There are many packages that perform this task andhere we use the [lisaClust function](https://www.bioconductor.org/packages/devel/bioc/html/lisaClust.html) that is based on "local L-function" to spatially cluster cells into different regions with similar cell type composition. 

```{r eval=params$evalc}
set.seed(51773)
BPPARAM <- MulticoreParam(16)

# Cluster cells into spatial regions with similar composition.
data_sce  <- lisaClust(
  data_sce ,
  k = 5,
  Rs = c(20, 50, 100),
  sigma = 50,
  spatialCoords = c("Location_Center_X", "Location_Center_Y"),
  cellType = "description",
  imageID = "ImageNumber" ,
  regionName = "region",
  BPPARAM = BPPARAM
)
```


## Which regions appear to be different between responders and non-responders?  

####  Visualise regions on individual level {-}

```{r fig.height=3, fig.width=10 , eval=params$evalc}
metadata <- colData(data_sce)
metadata <- metadata[metadata$metabricId == "MB-0263",  ]
metadata <- data.frame(metadata)

plotlist <- list()
plotlist_celltype <- list()
thisregion  <-  unique(metadata$region)[1]

tableau_palette <- scale_colour_tableau() 
color_codes <- tableau_palette$palette(10)
color_codes <- c(color_codes, "darkgrey") 

names(color_codes) <- c(unique(metadata$description) ,  "other regions")

for ( thisregion in sort(unique(metadata$region))){
        
        selected_region_index <-  metadata$region == thisregion
        
        metadata$selected_region <-  "other regions"
        metadata$selected_region[selected_region_index] <- "selected region"
        
        metadata$celltype <- metadata$description
        metadata$celltype[!selected_region_index ] <-   "other regions"
        
        metadata$celltype <- factor(metadata$celltype, levels = c(unique(metadata$description), "other regions"))

       p <- ggplot(metadata, aes(x = Location_Center_X , y = Location_Center_Y , colour = selected_region  )) + geom_scattermore(pointsize = 1.5) + ggtitle(thisregion) + scale_colour_manual(values = c("grey" , "red"))
         
       
    
       p2 <-  ggplot(metadata, aes(x = Location_Center_X , y = Location_Center_Y , colour =  celltype )) + geom_scattermore(pointsize = 1.5) + ggtitle(thisregion) + scale_colour_manual(values =  color_codes)
       
      plotlist [[thisregion ]] <- p
       
      plotlist_celltype [[thisregion ]] <- p2
}

ggarrange(plotlist = plotlist , ncol = 5, nrow = 1 , common.legend = T )
ggarrange(plotlist = plotlist_celltype , ncol = 5, nrow = 1 , common.legend = T )

```

####  Visualise regions across patients {-}

We can better interpret the region output by summarising the proportion of each cell type in a region across the patients. Looking at the composition of cell types in each region, comparing between Young and Old Patients. A cutoff of 50 years old will be used. 


```{r  fig.height=4, fig.width=10, eval=params$evalc}
df <- data.frame(colData( data_sce))
clinical$AgeGroup <- ifelse(clinical$Age.At.Diagnosis < 50, "Young", "Old")

df_plot <- NULL
for ( thispatient in unique(df$metabricId)){
  this_df <- df[df$metabricId == thispatient, ]
  temp_df <-   table( this_df$region, this_df$description )
  temp_df <-  temp_df / rowSums(temp_df)
   temp_df <- data.frame(  temp_df)
  temp_df$patient <-  thispatient
  temp_df$group <- clinical[clinical$metabricId == thispatient, "AgeGroup"]
  df_plot <- rbind(df_plot, temp_df)
}

df_plot <- df_plot %>% dplyr::group_by( Var1 , Var2, group) %>% 
  summarise(mean_proportion = mean(Freq))
  
# r1 <- df_plot[ df_plot$Var1 == "region_1", ]  

ggplot(df_plot, aes(y = Var2, x = Var1 ,colour =mean_proportion  , size = mean_proportion ))+  geom_point() + 
  facet_grid(~group, scales = "free", space = "free" ) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  xlab("Region" ) + ylab("Celltype") + scale_colour_viridis_c()
```


<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<i class="far fa-chart-bar fa-stack-1x fa-inverse"></i>
</span> <strong>Interactive Q&A:</strong>   
 
Q4: Which regions appear to be different between Young and Old?        

</div>


```{r  fig.height=4, fig.width=10, eval=params$evalc}
df <- clinical[colData(data_sce)[, "metabricId"], ]
df$region <- data_sce$region
df$description <- data_sce$description

df <- df %>% dplyr::group_by(metabricId , AgeGroup, region) %>%
  count(description) %>%
  mutate(proportion = n / sum(n))


ggplot(df, aes(y = proportion, x = metabricId , fill = description))+ geom_col()+facet_grid(~region+AgeGroup, scales = "free", space = "free" ) + scale_fill_manual(values = c(color_codes))  +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())


```


<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<i class="far fa-chart-bar fa-stack-1x fa-inverse"></i>
</span> <strong>Interactive Q&A:</strong>

Q5: Does your conclusion change after looking at a different plot?    

</div>


## Further exploration by visualising selected regions    
We see that region 1 appears to suggest the non-responder patients have more melano. Region 3 appears to be the tumor microenvironment with lots of Th.ae (antigen-experienced) and macro.mono (macrophage and monocytes) cell types. Let's focus on region 1 and region 3 and visualise the data with boxplots, as well as comparing to the overall cell type proportion without segmenting into regions.   

```{r fig.height=4, fig.width= 12 , eval=params$evalc}
df_plot <- NULL
for ( thispatient in unique(df$metabricId)){
  this_df <- df[df$metabricId == thispatient, ]
  temp_df <-   table( this_df$region, this_df$description )
  temp_df <-  temp_df / rowSums(temp_df)
  temp_df <- data.frame(  temp_df)
  temp_df$patient <-  thispatient
  temp_df$group <- unique( this_df$AgeGroup)
  df_plot <- rbind(df_plot, temp_df)
}

df_plot_region_1 <- df_plot[df_plot$Var1 == "region_1", ]
 
a <- ggplot(df_plot_region_1, aes(x =  Var2,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  ylab("Proportion") + xlab("Cell type")+ ggtitle("Region 1") + ylim(0, 0.25)


df_plot_region_3 <- df_plot[df_plot$Var1 == "region_3", ]

b <- ggplot(df_plot_region_3, aes(x =  Var2,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))  + 
  ylab("Proportion") + xlab("Cell type") + ggtitle("Region 3") + ylim(0, 0.25)
 


overall <- NULL
for ( thispatient in unique(df$metabricId)){
  
  this_df <- df[df$metabricId == thispatient, ]
  
  temp_df <-   table(  this_df$description )
  temp_df <-  temp_df /sum(temp_df)
   temp_df <- data.frame(  temp_df)
  temp_df$patient <-  thispatient
  temp_df$group <- unique( this_df$AgeGroup )
  overall <- rbind(overall, temp_df)
}


c <- ggplot(overall, aes(x =  Var1,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))  + 
  ylab("Proportion") + xlab("Cell type") + ggtitle("Overall composition")
 

a + b + c
```

#### Selecting a specific marker for further exploration {-}
Often you may have a marker in mind to further examine the expression of key marker genes in these region specific cell types.  For example, we select cells that have high Ki67 expression.  (ie, only keeping the cells that have Ki67 expression higher than the median Ki67 expression in the whole dataset). We choose Ki67 as an example here because Ki67 is strongly associated with tumor cell proliferation and growth and is widely used as a biomarker in cancer analysis.    


```{r fig.height=4, fig.width= 12, eval=params$evalc}

median_ki67 <- median( logcounts(data_sce[ "Ki67" , ]))
data_sce$ki67 <- ifelse( logcounts(data_sce[ "Ki67" , ]) > median_ki67, "high_ki67" , "low_ki67")


df_plot <- NULL
for ( thispatient in unique(df$metabricId)){
  this_df <- df[df$metabricId == thispatient, ]
  temp_df <-   table( this_df$region, this_df$description )
  temp_df <-  temp_df / rowSums(temp_df)
  temp_df <- data.frame(  temp_df)
  temp_df$patient <-  thispatient
  temp_df$group <- unique( this_df$AgeGroup )
  df_plot <- rbind(df_plot, temp_df)
}

df_plot_region_1 <- df_plot[df_plot$Var1 == "region_1", ]
 
a <- ggplot(df_plot_region_1, aes(x =  Var2,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  ylab("Proportion") + xlab("Cell type")+ ggtitle("Region 1") + ylim(0, 0.25)


df_plot_region_3 <- df_plot[df_plot$Var1 == "region_3", ]

b <- ggplot(df_plot_region_3, aes(x =  Var2,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))  + 
  ylab("Proportion") + xlab("Cell type") + ggtitle("Region 3") + ylim(0, 0.25)
 


overall <- NULL
for ( thispatient in unique(df$metabricId)){
  
  this_df <- df[df$metabricId == thispatient, ]
  
  temp_df <-   table(  this_df$description )
  temp_df <-  temp_df /sum(temp_df)
   temp_df <- data.frame(  temp_df)
  temp_df$patient <-  thispatient
  temp_df$group <- unique( this_df$AgeGroup )
  overall <- rbind(overall, temp_df)
}


c <- ggplot(overall, aes(x =  Var1,  y = Freq, colour = group)) +
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))  + 
  ylab("Proportion") + xlab("Cell type") + ggtitle("Overall composition") + ylim(0, 0.25)
 

a + b + c
```

<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<span class="far fa-comment fa-stack-1x fa-inverse"></span>
</span> <strong> Discussion:</strong>

Comparing the overall composition and the cell type composition in the region, what did you observe about the regions? 
</div>


#  How do we generate a molecular representation for each individual?  

In this demo, we use scFeatures to generate molecular representation for each patient from the matrix of proteins by cells. The molecular representation is interpretable and hence facilitates downstream analysis of the patient. Overall, scFeatures generates features across six categories representing different molecular views of cellular characteristics. These include:

i) cell type proportions
ii) cell type specific gene expressions
iii) cell type specific pathway expressions
iv) cell type specific cell-cell interaction (CCI) scores
v) overall aggregated gene expressions
vi) spatial metrics

The different types of features constructed enable a more comprehensive multi-view understanding of each patient from a matrix of proteins x cells.

Given in the previous section, we clustered the cells into regions, we can use the region information as an additional layer of information on top of the cell types to examine region-specific cell-type specific features. 

```{r eval=params$evalc}

region <- data_sce$region
region <- gsub("_" , "", region)
data_sce$celltype <- paste0( data_sce$description , "-" , region)

print("number of cells in each sample")
table(data_sce$metabricId) 

print("number of cells in each celltype - Region specific cell type")
table(data_sce$celltype) 
```

<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<span class="far fa-comment fa-stack-1x fa-inverse"></span>
</span> <strong> Discussion:</strong>

Are there any samples or cell types you would like to remove from the data?

</div>


## How to create molecular representations of patients {-}

All the feature types can be generated in one line of code. This runs the function using default settings for all parameters. For more information, type `?scFeatures`.


```{r eval=FALSE, include=TRUE, eval=params$evalc}

# scFeatures requires the following columns 
# celltype, sample, x_cord and y_cord
# alternatively, these can be also set as argument in the scFeatures function 
 
IMCmatrix <- assay(data_sce)
sample = data_sce$metabricId
celltype = data_sce$celltype
spatialCoords <- list(colData(data_sce)$Location_Center_X, colData(data_sce)$Location_Center_Y)

# here, we specify that this is a spatial proteomics data
# scFeatures support parallel computation to speed up the process 
scfeatures_result <- scFeatures(IMCmatrix, type = "spatial_p", sample = sample, celltype = celltype, spatialCoords = spatialCoords,
                                ncores = 32)
```


## How to explore scFeatures output?
We have generated a total of 13 feature types and stored them in a list. All generated feature types are stored in a matrix of samples by features.
For example, the first list element contains the feature type `proportion_raw`, which contains the cell type proportion features for each patient sample. We could print out the first 5 columns and first 5 rows of the first element to see.

```{r eval=params$evalc}
# scfeatures_result <- readRDS("scfeatures_result_region_specific.rds")

# combine sample name with outcome 
scfeatures_result_format <- scfeatures_result

# we have generated a total of 13 feature types
names(scfeatures_result_format)

# each row is a sample, each column is a feature 
data.frame(scfeatures_result_format[[1]][1:5, 1:5])
## DT::datatable(meta_table , options = list(scrollX = TRUE))
```

To accommodate for easier interpretation of the features, scFeatures contains a function `run_association_study_report` that enables the user to readily visualise and explore all generated features with one line of code.

**Tips:** Some categories of features such as cell cell interaction takes a long time to run.  You may use parallel computation to speed up the process or select a small number of feature categories to reduce computational time. 
```{r include=TRUE, eval=FALSE}
# specify a folder to store the html report. Here we store it in the current working directory. 
output_folder <-  getwd()
run_association_study_report(scfeatures_result, output_folder )
```

## Are the generated features sensible? 
<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<span class="far fa-comment fa-stack-1x fa-inverse"></span>
</span> <strong>Interactive Q&A:</strong>

Using the HTML, we can look at some of the critical thinking questions that a researcher would ask about the generated features. These questions are exploratory and there is no right or wrong answer.   

Q7: Do the generated features look reasonable?         

- Which cell type(s) would you like to focus on at your next stage of analysis?       
- Which feature type(s) would you like to focus on at your next stage of analysis?       
- Are the conditions in your data relatively easy or difficult to distinguish?       

</div>


# Can we predict recurrence risk? 

### Building a classification model {-}

In this section, we will perform disease outcome classification using the molecular representation of patients. Recall in the previous section that we have stored the 13 feature types matrix in a list. Instead of manually retrieving each matrix from the list to build separate models, classifyR can directly take a list of matrices as an input and run a repeated cross-validation model on each matrix individually. Below, we run 5 repeats of 5-fold cross-validation. The patient outcome is a Surv object, so, by default, ClassifyR will use Cox proportional hazards ranking to choose a set of features and also Cox proportional hazards to predict risk scores. Other models are available. A high score indicates prection of a worse outcome than a lower risk score.

```{r eval=FALSE, include=TRUE}
outcome <- clinical[rownames(scfeatures_result[[1]]), c("timeRFS", "eventRFS")]

usefulFeatures <- c("Breast.Tumour.Laterality", "ER.Status", "Inferred.Menopausal.State", "Grade", "Size")
scfeatures_result <- append(list(clinical = clinical[rownames(scfeatures_result[[1]]), usefulFeatures]), scfeatures_result)
nFeatures <- append(list(clinical = 1:3), lapply(scfeatures_result[2:length(scfeatures_result)], function(metaFeature) 1:5))

### generate classfyr result 

classifyr_result <- crossValidate(scfeatures_result, outcome, nFeatures = nFeatures, nFolds = 5, nRepeats = 5,  nCores = 5)
```

### Visualising the classification performance of individual metafeatures {-}
To examine the distribution of prediction performance. use `performancePlot`. Currently, the only metric is C-index and that will automatically be used because the preditive model type is tracked inside of the results objects.

```{r eval=params$evalc}
tilt <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
performancePlot(classifyr_result, orderingList = list(`Assay Name` = names(scfeatures_result))) + tilt
```

Note how the resultant plot is a ggplot2 object and can be further modified.

Next, examine feature selection stability with `selectionPlot`.

```{r}
stabilityPlot <- selectionPlot(classifyr_result, orderingList = list(`Assay Name` = names(scfeatures_result)), plot = FALSE)
stabilityPlot + tilt
distribution(classifyr_result[[1]], plot = FALSE)
```

<div class="aimbox"> 
<span class="fa-stack fa-lg">
<i class="fa fa-circle fa-stack-2x"></i>
<i class="far fa-chart-bar fa-stack-1x fa-inverse"></i>
</span> <strong>Interactive Q&A:</strong>   

Q8: Is highest predictive performance the only way to choose the best model or are other models the best?
</div>

## PART A: Explanation of spatial features 

- L function:

The L function is a spatial statistic used to assess the spatial distribution of cell types. It assesses the significance of cell-cell interactions, by calculating the density of a cell type with other cell types within a certain radius. High values indicate spatial association, low values indicate spatial avoidance.

```{r fig.height=4, fig.width=10, eval=params$evalc}

tableau_palette <- scale_colour_tableau() 
color_codes <- tableau_palette$palette( 10 )
color_codes <- c(color_codes, "darkgrey" , "grey90") 

names(color_codes) <- c( unique(data_sce$description) ,  "other regions")
 
one_sample  <- data_sce[ , data_sce$metabricId == "16BL"  ]
one_sample <- data.frame( colData(one_sample) )

one_sample$celltype <- one_sample$description
index <-  one_sample$celltype  %in% c("macro.mono", "Tc.ae")
one_sample$celltype[!index] <- "others"
a <-ggplot( one_sample, aes(x = Location_Center_X , y = Location_Center_Y, colour = celltype )) + geom_point()  + scale_colour_manual(values = color_codes)  + ggtitle( "Patient 16BL - high L value with \n macro.mono interacting Tc.ae")
 

one_sample$celltype <- one_sample$description
index <-  one_sample$celltype  %in% c("melano", "Tc.ae")
one_sample$celltype[!index] <- "others"
b <- ggplot( one_sample, aes(x = Location_Center_X , y = Location_Center_Y, colour = celltype )) + geom_point()  + scale_colour_manual(values = color_codes)  + ggtitle( "Patient 16BL - low L value with  \n melano interacting Tc.ae")
 
a + b

 
```




- Cell type interaction composition:

We calculate the nearest neighbours of each cell and then calculate the pairs of cell types based on the nearest neighbour. This allows us to summarise it into a cell type interaction composition.


```{r fig.height=6, fig.width=10, eval=params$evalc}

tableau_palette <- scale_colour_tableau() 
color_codes <- tableau_palette$palette( 10 )
color_codes <- c(color_codes, "darkgrey" , "grey90") 

names(color_codes) <- c( unique(data_sce$description) ,  "other regions")
 
one_sample  <- data_sce[ , data_sce$metabricId == "16BL"  ]
one_sample <- data.frame( colData(one_sample) )

one_sample$celltype <- one_sample$description
 
a <-ggplot( one_sample, aes(x = Location_Center_X , y = Location_Center_Y, colour = celltype )) + geom_point()  + scale_colour_manual(values = color_codes)  + ggtitle( "Patient 16BL")


feature  <- scfeatures_result$celltype_interaction
to_plot <- data.frame( t( feature["16BL", ])  )
to_plot$feature <- rownames(to_plot) 
colnames(to_plot)[2] <- "celltype interaction composition"
 
to_plot <- to_plot[ to_plot$X16BL > 0.03 , ] 
b <- ggplot(to_plot, aes( x =  reorder(`celltype interaction composition`, X16BL) ,  y = X16BL, fill=`celltype interaction composition`)) + geom_bar(stat="identity" ) + ylab("Major cell type interactions")  +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 

a+ b
```



- Moran's I:   

Moran's I is a spatial autocorrelation statistic based on both location and values. It quantifies whether similar values tend to occur near each other or are dispersed.     


```{r fig.height=4, fig.width=10 , eval=params$evalc}

high  <- data_sce["Ki67", data_sce$metabricId == "25RD"  ]
high_meta <- data.frame( colData(high) ) 
high_meta$expression <- as.vector(logcounts( high)) 

low  <- data_sce["Ki67",  data_sce$metabricId == "42RD" ]
low_meta <- data.frame( colData(low) )
low_meta$expression <- as.vector(logcounts(low))


a <- ggplot(high_meta, aes(x = Location_Center_X , y = Location_Center_Y, colour =expression)) + geom_point(alpha=0.5) + scale_colour_viridis_c() + ggtitle("Patient 25RD - high Moran's I in Ki67")

b <- ggplot(low_meta, aes(x = Location_Center_X , y = Location_Center_Y, colour =expression)) + geom_point(alpha=0.5) + scale_colour_viridis_c() + ggtitle("Patient 42RD - low Moran's I in Ki67")

a+b

```



- Nearest Neighbor Correlation:    

This metric measures the correlation of proteins/genes between a cell and its nearest neighbour cell.   

We see from both prediction and survival analysis that the feature type "nn correlation" (nearest neighbour correlation) performs the best.   


Here we pick the protein "S100", a key player in cancer, as an example to illustrate the concept.  

```{r fig.height=5, fig.width=10, eval=params$evalc}


    
 plot_nncorrelation <- function(thissample , thisprotein){
   
       sample_name <- thissample
       thissample <- data_sce[, data_sce$metabricId ==     sample_name]
    
      
      exprsMat <- logcounts(thissample)
     
    
    cell_points_cts <- spatstat.geom::ppp(
            x = as.numeric(thissample$Location_Center_X ), y = as.numeric(thissample$Location_Center_Y),
            check = FALSE,
            xrange = c(
                min(as.numeric(thissample$Location_Center_X)),
                max(as.numeric(thissample$Location_Center_X))
            ),
            yrange = c(
                min(as.numeric(thissample$Location_Center_Y)),
                max(as.numeric(thissample$Location_Center_Y))
            ),
            marks = t(as.matrix(exprsMat))
        )
    
     value <-  spatstat.explore::nncorr(cell_points_cts)["correlation", ]
      value <-  value[  thisprotein]
     
    # Find the indices of the two nearest neighbors for each cell
    nn_indices <- nnwhich(cell_points_cts, k = 1)
    
    protein <-  thisprotein
    df <- data.frame(thiscell_exprs  = exprsMat[protein, ] , exprs =  exprsMat[protein,nn_indices ])
    
   p <-  ggplot(df, aes( x =thiscell_exprs ,  y = exprs , colour =  exprs  )) +
      geom_point(alpha = 0.3) + ggtitle(paste0( "Patient ", sample_name ,  " nn_corr = " ,  round(value, 2)  )) + scale_colour_viridis_c() 
   
   return (p ) 

}

    
p1 <- plot_nncorrelation( "42RD" ,  "S100" )
p2 <- plot_nncorrelation( "29RD" ,  "S100" )
p1  + p2   
 
```

The correlation differs between the 42RD patient (from cluster 1) and the 29RD patient (from cluster 2).    


## PART C: SessionInfo

```{r  eval=params$evalc}
sessionInfo()
```

## Acknowledgment 
The authors thank all their colleagues, particularly at The University of Sydney, Sydney Precision Data Science and Charles Perkins Centre for their support and intellectual engagement. Special thanks to Ellis Patrick, Shila Ghazanfar, Andy Tran for guiding and supporting the building of this workshop.